alluvial diagrams in ggplot2

### 2017-11-06 inst/doc/ggalluvial.rmd

### One row per alluvium

The first format reflects the visual arrangement of an alluvial diagram, but "untwisted": Each row corresponds to a subset or amount of observations that take a specific value at each variable, and each variable has its own column. An additional column contains the weight of each row. This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:

```{r Berkeley admissions dataset}
head(as.data.frame(UCBAdmissions), n = 12)
is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE, silent = TRUE)
```

This form was inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**. It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:

@@

### One row per lode

The second format recognized by **ggalluvial** contains one row per lode, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset of the preceding kind into a "key" column indicating the axis and a "value" column indicating the value taken thereat. This format requires an additional indexing column that links the rows corresponding to a common alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:

```{r lode form of Berkeley admissions dataset}
UCB_lodes <- to_lodes(as.data.frame(UCBAdmissions), axes = 1:3)
head(UCB_lodes, n = 12)
is_alluvial(UCB_lodes, logical = FALSE, silent = TRUE)
```

@@

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes hierarchically:

@@

```{r time series alluvia diagram of refugees dataset}
data(Refugees, package = "alluvial")
country_regions <- c(
  Afghanistan = "Middle East",
  Burundi = "Central Africa",
  `Congo DRC` = "Central Africa",
  Iraq = "Middle East",
  Myanmar = "Southeast Asia",
  Palestine = "Middle East",
  Somalia = "Horn of Africa",
  Sudan = "Central Africa",
  Syria = "Middle East",
  Vietnam = "Southeast Asia"
)
Refugees$region <- country_regions[Refugees$country]
ggplot(data = Refugees,
       aes(x = year, weight = refugees, alluvium = country)) +
  geom_alluvium(aes(fill = country, colour = country),
                alpha = .75, decreasing = FALSE) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  scale_color_brewer(type = "qual", palette = "Set3") +
  facet_wrap(~ region, scales = "fixed") +
  ggtitle("refugee volume by country of origin and geographic region")
```

@@

```{r alluvial diagram of majors dataset}
data(majors)
majors$curriculum <- as.factor(majors$curriculum)
ggplot(majors,
       aes(x = semester, stratum = curriculum, alluvium = student,
           fill = curriculum, label = curriculum)) +
  geom_flow(stat = "alluvium", lode.guidance = "rightleft",
            color = "black") +
  geom_stratum() +
  theme(legend.position = "bottom") +
  ggtitle("student curricula across several semesters")
```

@@

This diagram ignores any continuity between the alluvial segments linking different pairs of adjacent axes. This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each cohort across the entire diagram.

>>>>

### Alluvia format

The first format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g. the number of observational units in the cohort. This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:

```{r alluvia format of Berkeley admissions dataset}
head(as.data.frame(UCBAdmissions), n = 12)
is_alluvial(as.data.frame(UCBAdmissions), logical = FALSE, silent = TRUE)
```

This format is inherited from the first version of **ggalluvial**, which modeled it after usage in **alluvial**. It required a stark departure from the usual position aesthetics: The user declares any number of axis variables, which `stat_alluvium()` and `stat_stratum()` recognize and process in a consistent way:

@@

### Lodes format

The second format recognized by **ggalluvial** contains _one row per lode_, and can be understood as the result of "gathering" (in the **dplyr** sense) the axis columns of a dataset in the alluvia format into a key-value pair of columns encoding the axis as the `key` and the stratum as the `value`. This format requires an additional indexing `id` column that links the rows corresponding to a common cohort, i.e. the lodes of a single alluvium, as illustrated below using the `to_lodes()` defaults on the Berkeley admissions dataset:

```{r lodes format of Berkeley admissions dataset}
UCB_lodes <- to_lodes(as.data.frame(UCBAdmissions), axes = 1:3)
head(UCB_lodes, n = 12)
is_alluvial(UCB_lodes, logical = FALSE, silent = TRUE)
```

@@

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and usually not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can grop countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:

@@

```{r time series alluvia diagram of refugees dataset}
data(Refugees, package = "alluvial")
country_regions <- c(
  Afghanistan = "Middle East",
  Burundi = "Central Africa",
  `Congo DRC` = "Central Africa",
  Iraq = "Middle East",
  Myanmar = "Southeast Asia",
  Palestine = "Middle East",
  Somalia = "Horn of Africa",
  Sudan = "Central Africa",
  Syria = "Middle East",
  Vietnam = "Southeast Asia"
)
Refugees$region <- country_regions[Refugees$country]
ggplot(data = Refugees,
       aes(x = year, weight = refugees, alluvium = country)) +
  geom_alluvium(aes(fill = country, colour = country),
                alpha = .75, decreasing = FALSE) +
  scale_x_continuous(breaks = seq(2003, 2013, 2)) +
  theme(axis.text.x = element_text(angle = -30, hjust = 0)) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  scale_color_brewer(type = "qual", palette = "Set3") +
  facet_wrap(~ region, scales = "fixed") +
  ggtitle("refugee volume by country and region of origin")
```

@@

```{r alluvial diagram of majors dataset}
data(majors)
majors$curriculum <- as.factor(majors$curriculum)
ggplot(majors,
       aes(x = semester, stratum = curriculum, alluvium = student,
           fill = curriculum, label = curriculum)) +
  geom_flow(stat = "alluvium", lode.guidance = "rightleft",
            color = "darkgray") +
  geom_stratum() +
  theme(legend.position = "bottom") +
  ggtitle("student curricula across several semesters")
```

@@

This diagram ignores any continuity between the flows between axes. This "memoryless" plot produces a less cluttered diagram, in which at most one flow proceeds from each stratum at one axis to each stratum at the next, but at the cost of being able to track each cohort across the entire diagram.
